A full approximation storage multilevel method \\ for nonlinear variational inequalities

### 2023-06-28 paper/fascd.tex
section 5 refinished; no FIXMEs outside of Discussion section at end

In Sections \ref{sec:femultilevel}--\ref{sec:vcycle} we will introduce multilevel CD discretizations and algorithms for finite-dimensional VI problems.  However, one can observe here that the CD concept applies at the level of the continuum problem, as illustrated in the following two examples, the first of which is an overlapping domain decomposition of a Sobolev space.

@@

Our second example is a disjoint frequency decomposition of a Hilbert space.  A multilevel CD over a finite-dimensional FE discretization, as proposed in Sections \ref{sec:femultilevel}--\ref{sec:cdmultilevel}, approximates such a decomposition.

@@

Note that $\cK_i \not\subset \cK$ in these cases; informally speaking, the important inclusion is $\cK_i \subset \cV_i$.

Associated to a constraint decomposition \eqref{eq:subspacedecomp}--\eqref{eq:constraintrestrictionsum} are certain iterative methods \cite{Tai2003,Xu1992}.  Let $0\le \alpha \le 1$ be the \emph{damping parameter}.  Algorithms \ref{alg:basiccd-add} and \ref{alg:basiccd-mult} below, discussed next, solve smaller VI problems over each subset $\cK_i$.  Each Algorithm replaces $u \in \cK$ with a new iterate $w\in\cK$ which should be closer to $u^* \in \cK$ solving \eqref{eq:vi}.

@@

In VIs \eqref{eq:cdaddvi} and \eqref{eq:cdmultvi} the argument of $f$ is an element of $\cK$, as the reader should confirm.

By \eqref{eq:constraintdecomp} and \eqref{eq:constraintrestrictionsum} one may write the test vector $v_i - \hat w_i \in \cV_i$ as a difference of admissible vectors from $\cK$.  For \eqref{eq:cdaddvi} and \eqref{eq:cdmultvi} respectively:
\begin{align}
[u - \Pi_i u + v_i] - [u - \Pi_i u + \hat w_i] &= v_i - \hat w_i, \label{eq:admissibledifferenceadd} \\
\left[\sum_{j<i} w_j + v_i + \sum_{j>i} \Pi_j u\right] - \left[\sum_{j<i} w_j + \hat w_i + \sum_{j>i} \Pi_j u\right] &= v_i - \hat w_i.  \label{eq:admissibledifferencemult}
\end{align}
In this sense the VIs over the subsets $\mathcal{K}_i$ solve the same problem as \eqref{eq:vi}.  Furthermore, let $e_i = \hat w_i - \Pi_i u \in \cV_i$.  The reader may confirm that $\hat w = u + \sum_{i} e_i$ and thus that $w = u + \alpha \sum_i e_i$.  The new iterate $w$ is therefore computed by adding a damped correction from each sub\emph{space} $\mathcal{V}_i$, but in such a manner that $w \in \mathcal{K}$; the corrections preserve admissibility.

@@

The basic CD iterations are meaningful whether or not we are in the optimization case $f=F'$, and when $f$ is nonlinear, non-local, or defined only on $\cK$.  However, practical and efficient FE implementation for such general operators $f$ seems not to have been addressed in the literature.  References \cite{GraeserKornhuber2009,Tai2003} only apply the CD method to the classical obstacle problem, and which case $f=F'$ is linear, local, and defined for all inputs (on all of $\mathcal{V}$).  Substantial implementation simplifications are available in such cases.

@@

When $f$ has the locality of Example \ref{ex:fnice}, an FE approximation of VI \eqref{eq:vi} can be solved in any incremental and efficient manner following the techniques in \cite{GraeserKornhuber2009,Tai2003}, namely by computations over hat function supports.\footnote{Compare equations (2.31) and (2.32) in \cite{Farrelletal2021}, which describe the same sense of locality.}  However, for general problems lacking the niceties of Example \ref{ex:fnice}, certain additional ideas are needed for practical implementation of \pr{cd-add} or \pr{cd-mult}.  Section \ref{sec:vcycle} constructs a multiplicative multilevel CD iteration for a general operator $f:\cK\to\cV'$ by applying the full approximation storage (FAS) idea of Brandt \cite{Brandt1977}.

From now on we restrict to the case of \emph{box constraints} \cite{BensonMunson2006,FerrisPang1997}, equivalently \emph{bi-obstacle problems}.  Suppose $\Omega \subset \RR^d$ is open and let $\mathcal{V}=W^{1,p}(\Omega)$.  The \emph{lower obstacle} $\underline{\gamma}$ and the \emph{upper obstacle} $\overline{\gamma}$ are given functions, defined on the closure $\overline{\Omega}$, with values in the extended real line $\tilde\RR = [-\infty,+\infty]$.  Specifically, we suppose $\underline{\gamma} : \overline{\Omega} \to [-\infty,+\infty)$ and $\overline{\gamma} : \overline{\Omega} \to (-\infty,+\infty]$ are measurable, $\underline{\gamma} \le \overline{\gamma}$ on $\overline{\Omega}$, and $\underline{\gamma} \le g_D \le \overline{\gamma}$ on $\partial_D \Omega$.  Also, we assume that $\partial\Omega$ is split into disjoint, measurable sets, $\partial\Omega = \partial_D \Omega \cup \partial_N \Omega$, with $\partial_D \Omega$ of positive measure, to facilitate well-posedness \cite{Evans2010}, and $\partial_N \Omega$ possibly empty.  Dirichlet data\footnote{Dirichlet boundary values are in the trace sense \cite{Evans2010}, and $g_D$ is assumed to be as regular as needed.} $g_D:\partial_D \Omega \to \RR$ then defines
\begin{equation}
\cK = \left\{w\,:\,\underline{\gamma} \le w \le \overline{\gamma} \, \text{ and }\, w\big|_{\partial_D \Omega} = g_D\right\} \subset \cV =W^{1,p}(\Omega), \label{eq:originalconstraintset}
\end{equation}
the closed and convex constraint set.  Note that we are treating inequality and equality (boundary) constraints in a unified manner.

Recalling abstract VI problem \eqref{eq:vi}, we now consider box-constrained cases only, using set \eqref{eq:originalconstraintset}.  Given continuous $f:\cK \to \cV'$, and $\ell \in \cV'$, we will assume that the following is well-posed for $u^*\in \cK$:
\begin{equation}
\ip{f(u^*)}{v-u^*} \ge \ip{\ell}{v-u^*} \qquad \text{for all } v\in \cK. \label{eq:boxdirichletvi}
\end{equation}
Note that $v-u^* \in W_0^{1,p}(\Omega)$, so test functions in \eqref{eq:boxdirichletvi} have zero Dirichlet values.  Any Neumann conditions over $\partial_N \Omega$ are determined by the operator formula for $f$, i.e.~by an integration over $\partial_N\Omega$.  In summary, from now on we consider well-posed VI problems \eqref{eq:boxdirichletvi}, namely box-constrained boundary value problems.

@@

We seek to rapidly compute the approximate solution of box-constrained VI problems using finite elements (FE).  Suppose $\Omega \subset \RR^d$ is a (open) bounded polygon.\footnote{Section \ref{sec:results} illustrates $d=1,2,3$ cases.}  We will construct a multilevel FE approximation to problem \eqref{eq:boxdirichletvi} over $\Omega$ using $P_1$ or $Q_1$ elements \cite{Elmanetal2014}, by refining a coarsest mesh into a mesh hierarchy; the goal is an accurate solution on the finest-level mesh.  The hierarchy allows the creation of a multilevel CD of $\cK$, using the ideas in the next two Sections, which permits admissibility-preserving coarse corrections.

@@

In the VI material which follows, use of higher-order Lagrange elements, i.e.~$P_k$ or $Q_k$ with $k\ge 2$, would imply a ``variational crime'' \cite[Chapter 10]{BrennerScott2007}, special to inequality constraints, which we avoid.  Namely, our FE construction of constraint sets (Section \ref{sec:cdmultilevel}) will require \emph{nodal monotonicity},
\begin{equation}
\bw \ge \bz \quad \implies \quad w \ge z \label{eq:nodalmonotonicity}
\end{equation}
for all $w,z \in \mathcal{V}^j$, where $\bw,\bz \in \RR^{m_j}$ denote vectors of nodal values.  However, for $k\ge 2$ there are hat functions in $P_k$ and $Q_k$, i.e.~functions with nodal values one or zero, which take on negative values \cite[Figure 1.7]{Elmanetal2014}.  Evidently \eqref{eq:nodalmonotonicity} is equivalent to the non-existence of such indefinite-sign hat functions.

To allow box constraints on all meshes in the hierarchy, let $\tilde{\mathcal{V}}^j$ denote the set of functions\footnote{Note that $\tilde{\mathcal{V}}^j$ functions need not be defined everywhere on $\Omega$.} on the nodes of $\mathcal{T}^j$ with values in the extended real line $\tilde{\RR} = [-\infty,+\infty]$.  Suppose that the finest-level \emph{lower} and \emph{upper obstacles} $\underline{\gamma}^J, \overline{\gamma}^J \in \tilde{\mathcal{V}}^J$ satisfy
\begin{equation}
\underline{\gamma}^J \le \overline{\gamma}^J, \qquad \underline{\gamma}^J < +\infty, \qquad -\infty < \overline{\gamma}^J. \label{eq:fe:boxconstraintrequirements}
\end{equation}
Also assume that $\underline{\gamma}^J \le g_D^J \le \overline{\gamma}^J$ on those nodes of $\mathcal{T}^J$ which lie in $\partial_D \Omega$.  Define
\begin{equation}
\mathcal{K}^J = \left\{w\,:\,\underline{\gamma}^J \le w \le \overline{\gamma}^J \text{ on nodes of } \mathcal{T}^J, \, \text{ and } \, w|_{\partial_D\Omega} = g_D^J|_{\partial_D\Omega}\right\} \subset \mathcal{V}^J, \label{eq:fe:fineconstraintset}
\end{equation}
the finest-level \emph{constraint set} of admissible FE functions.  This is a nonempty, closed, and convex subset of $\mathcal{V}^J$.  As $\mathcal{V}^J$ functions never take values $\pm\infty$, inequality in \eqref{eq:fe:fineconstraintset} is strict where $\underline{\gamma}^J, \overline{\gamma}^J$ are infinite; there is no lower or upper constraint, respectively, at such nodes.

@@

While \eqref{eq:fe:vi} approximates \eqref{eq:boxdirichletvi}, we warn the reader that it may not technically be conforming for that VI.  That is, even if quadrature and other implementation details for $f^J$ are ignored, and despite the vector space inclusion $\mathcal{V}^J \subset \mathcal{V}$, we do not require or generally expect $\mathcal{K}^J \subset \mathcal{K}$.  Regardless of the smoothness of the continuum obstacles $\underline{\gamma}, \overline{\gamma}$, admissibility with respect to their FE interpolants $\underline{\gamma}^J, \overline{\gamma}^J$ will generally not imply continuum admissiblity.\footnote{This caveat is especially relevant in geophysical situations where only nodal values of the obstacle are known.  The nodal elevation data misses the actual topographic detail in-between the nodes.  See the Greenland ice sheet application in \cite{Bueler2016}, for example.}

The multilevel algorithm defined in the next two Sections, a solver for FE VI \eqref{eq:fe:vi}, requires transfer operators between meshes.  We require five such operators, of which three are standard multigrid tools \cite{Trottenbergetal2001}.  Table \ref{tab:transfers} will help the reader keep track.

@@

Canonical prolongation $P:\mathcal{V}^{j-1}\to\mathcal{V}^j$ is the vector space inclusion $\mathcal{V}^{j-1} \hookrightarrow \mathcal{V}^j$ of coarser-mesh functions.  Canonical (dual) restriction $R:(\mathcal{V}^j)'\to(\mathcal{V}^{j-1})'$ also acts as an identity in the sense that if $\sigma \in (\mathcal{V}^j)'$ then $(R\sigma)[z] = \sigma[z]$ for all $z \in \mathcal{V}^{j-1} \subset \mathcal{V}^j$.  Note that these canonical transfers generally involve a change of FE representation.  For example, the formula for canonical restriction, derived by writing coarser-mesh hat functions in terms of finer-mesh ones, is ``full-weighting'' \cite{Trottenbergetal2001}.

To define injection $\iR:\mathcal{V}^j\to\mathcal{V}^{j-1}$, observe that the nodes of $\mathcal{T}^{j-1}$ are nodes of $\mathcal{T}^j$, and that for any values at $\mathcal{T}^j$ nodes there is a unique $\mathcal{V}^j$ function assuming those values.  For $z\in\mathcal{V}^j$ we define the injection $\iR z$ as the element of $\mathcal{V}^{j-1}$ with the same (coarse) nodal values.  Note that injection is monotonic: if $w \le z$ then $\iR w \le \iR z$.

Finally, the monotone restrictions \cite{GraeserKornhuber2009} $\maxR,\minR$ are nonlinear maps which maximize and minimize, respectively, nodal values over the open supports of coarser-mesh hat functions.  These monotone operators will only be applied to nodal functions $\tilde{\mathcal{V}}^j$.  To make the definitions precise, let $x_q^j$, $q=0,1,\dots,m_j-1$, denote the nodes of mesh $\mathcal{T}^j$, and denote the corresponding hat functions by $\psi_q^j(x)$.  The monotone restrictions of an $\tilde\RR$-valued\footnote{In computing $\max$ and $\min$ to define \eqref{eq:fe:monotonerestrictions}, the usual $\tilde\RR$ ordering is applied.} nodal function $z\in\tilde{\mathcal{V}}^j$ are the following functions in $\tilde{\mathcal{V}}^{j-1}$:
\begin{align}
(\maxR z)(x_p^{j-1}) &= \max \{z(x_q^j) \,:\, \psi_p^{j-1}(x_q^j) > 0\}, \label{eq:fe:monotonerestrictions} \\
(\minR z)(x_p^{j-1}) &= \min \{z(x_q^j) \,:\, \psi_p^{j-1}(x_q^j) > 0\}, \notag
\end{align}
for $p=0,1,\dots,m_{j-1}-1$.  By standard nested refinement, a coarse-mesh node $x_p^{j-1}$ is in the set of fine nodes $x_q^j$ such that $\psi_p^{j-1}(x_q^j)>0$, so it follows that
\begin{equation}
\minR z \le z \le \maxR z,  \label{eq:fe:monotonerestrictionprops}
\end{equation}
On the other hand, $z\ge 0$ implies $\minR z \ge 0$ and $z \le 0$ implies $\maxR z \le 0$.  These ordering properties used in the CD construction of the next two sections.

@@

Our practical multilevel CD scheme in this Section uses \emph{defect constraints} \cite{GraeserKornhuber2009} which are based on an admissible finest-level iterate.  The next Example concretely illustrates why we do not apply restriction operators directly to the finest-level box constraints $\underline{\gamma}^J,\overline{\gamma}^J$ in definition \eqref{eq:fe:fineconstraintset}.  Our straightforward iterate-dependent bounds, namely defect constraints \eqref{eq:fe:defectconstraints} below, bypass the difficulties exposed in Example \ref{ex:directRbad}.  However, other successful schemes may exist.

\begin{example}  \label{ex:directRbad}  
For concreteness, suppose $\Omega = [0,L] \subset \RR^1$ is an interval and consider a two-level mesh hierarchy ($J=1$).  Let $\underline{\gamma}^1$ be a sawtooth function with minimum value zero, and suppose $\overline{\gamma}^1=C+\underline{\gamma}^1$ with $C>0$ which is strictly less than the range of $\underline{\gamma}^1$.  Supposing empty Dirichlet boundary for simplicity, the constraint set $\mathcal{K}^1$ defined by \eqref{eq:fe:fineconstraintset} is nonempty.  Figure \ref{fig:directRbad} shows such functions; here levels $\mathcal{T}^1$ and $\mathcal{T}^0$ have $5$ and $3$ nodes, respectively.  However, various restriction schemes applied directly to $\underline{\gamma}^1,\overline{\gamma}^1$ are problematical.  Suppose we propose
    $$\underline{\gamma}^0, \overline{\gamma}^0 \equiv \iR \underline{\gamma}^1, \iR \overline{\gamma}^1. \qquad \text{\emph{[problematic]}}$$
Then $\underline{\gamma}^0=0$ and $\overline{\gamma}^0=C$ identically.  In that case the corresponding constraint set $\mathcal{K}^0$ is nonempty, but prolongation $Pw^0$ of an admissible solution of the coarse problem ($\underline{\gamma}^0 \le w^0 \le \overline{\gamma}^0$) is not admissible on the finer level: $Pw^0 \notin \mathcal{K}^1$.  On the other hand, suppose we propose
    $$\underline{\gamma}^0, \overline{\gamma}^0 \equiv \maxR \underline{\gamma}^1, \minR \overline{\gamma}^1. \qquad \text{\emph{[problematic]}}$$
The general motivation for this choice---compare \eqref{eq:fe:chilevels}---is that $w^0 \ge \underline{\gamma}^0 \, \implies \, Pw^0 \ge \underline{\gamma}^1$ and $w^0 \le \overline{\gamma}^0 \, \implies \, Pw^0 \le \overline{\gamma}^1$.  However, while these facts are true, the difficulty should be clear: the coarse-level admissible set $\mathcal{K}^0$ is now empty because the proposed lower bound $\underline{\gamma}^0$ actually exceeds the proposed upper bound $\overline{\gamma}^0$.
\end{example}

@@

Subtraction of an admissible iterate provides a route out of these difficulties.  Suppose $w^J \in \cK^J$ is an admissible finest-level iterate (i.e.~which approximates the solution $u^J$ of \eqref{eq:fe:vi}).  We define the \emph{finest-level defect constraints} (DCs) \cite{GraeserKornhuber2009} as the following nodal functions in $\tilde{\mathcal{V}}^J$:
\begin{equation}
\underline{\chi}^J = \underline{\gamma}^J - w^J, \qquad \overline{\chi}^J = \overline{\gamma}^J - w^J. \label{eq:fe:defectconstraints}
\end{equation}
Definition \eqref{eq:fe:fineconstraintset} implies that $\underline{\chi}^J \le 0 \le \overline{\chi}^J$ everywhere in $\overline{\Omega}$, including along the Dirichlet boundary $\partial_D\Omega$.  If $\underline{\gamma}^J=-\infty$ then $\underline{\chi}^J=-\infty$ by definition, and similarly if $\overline{\gamma}^J=+\infty$ then $\overline{\chi}^J=+\infty$.  Observe that a corrected iterate $w^J + z^J$ is admissible, $w^J + z^J \in \cK^J$, if and only if the perturbation $z^J$ is zero on $\partial_D\Omega$ and it is bounded by the defect constraints, $\underline{\chi}^J \le z^J \le \overline{\chi}^J$.

>>>>

In Sections \ref{sec:femultilevel}--\ref{sec:vcycle} we will introduce multilevel CD discretizations and algorithms for finite-dimensional VI problems.  However, one can observe here that the CD concept applies at the level of the continuum problem, as illustrated in the following two examples, the first of which is an overlapping domain decomposition of a Sobolev space.  Note that $\cK_i \not\subset \cK$ in these cases; informally speaking, the important inclusion is $\cK_i \subset \cV_i$.

@@

Our second example is a disjoint frequency decomposition of a Hilbert space.  A multilevel CD over a finite-dimensional FE discretization, as proposed in Sections \ref{sec:femultilevel} and \ref{sec:cdmultilevel}, approximates such a decomposition.

@@

Associated to a constraint decomposition \eqref{eq:subspacedecomp}--\eqref{eq:constraintrestrictionsum} are certain iterative methods \cite{Tai2003,Xu1992}.  Let $0\le \alpha \le 1$ be the \emph{damping parameter}.\footnote{FASCD Algorithms \ref{alg:fascd} and \ref{alg:fascd-fcycle} are presented without damping, and indeed it seems to be unnecessary.  However, the general convergence of the basic CD iterations appears to depend upon it.}  Algorithms \ref{alg:basiccd-add} and \ref{alg:basiccd-mult} below, discussed next, solve smaller VI problems over each subset $\cK_i$.  Each Algorithm replaces $u \in \cK$ with a new iterate $w\in\cK$ which should be closer to $u^* \in \cK$ solving \eqref{eq:vi}.

@@

In VIs \eqref{eq:cdaddvi} and \eqref{eq:cdmultvi} the argument of $f$ is an element of $\cK$, as the reader should confirm.  Furthermore, by \eqref{eq:constraintdecomp} and \eqref{eq:constraintrestrictionsum} one may write the test vector $v_i - \hat w_i \in \cV_i$ as a difference of admissible vectors from $\cK$.  For \eqref{eq:cdaddvi} and \eqref{eq:cdmultvi} respectively:
\begin{align}
[u - \Pi_i u + v_i] - [u - \Pi_i u + \hat w_i] &= v_i - \hat w_i, \label{eq:admissibledifferenceadd} \\
\left[\sum_{j<i} w_j + v_i + \sum_{j>i} \Pi_j u\right] - \left[\sum_{j<i} w_j + \hat w_i + \sum_{j>i} \Pi_j u\right] &= v_i - \hat w_i.  \label{eq:admissibledifferencemult}
\end{align}
In this sense the VIs over the subsets $\mathcal{K}_i$ solve the same problem as \eqref{eq:vi}.

Let $e_i = \hat w_i - \Pi_i u \in \cV_i$.  The reader may confirm that $\hat w = u + \sum_{i} e_i$ and thus that $w = u + \alpha \sum_i e_i$.  The new iterate $w$ is therefore computed by adding a damped correction from each sub\emph{space} $\mathcal{V}_i$, but in such a manner that $w \in \mathcal{K}$; the corrections preserve admissibility.

@@

The basic CD iterations are meaningful whether or not we are in the optimization case $f=F'$, and even when $f$ is nonlinear, non-local, or defined only on $\cK$.  However, practical and efficient FE implementation for such general operators $f$ seems not to have been addressed in the literature.  References \cite{GraeserKornhuber2009,Tai2003} only apply the CD method to the classical obstacle problem, wherein $f=F'$ is linear, local, and defined for all inputs (on all of $\mathcal{V}$).  Substantial implementation simplifications are available in such cases.

@@

When $f$ has the locality of Example \ref{ex:fnice}, an FE approximation of VI \eqref{eq:vi} can be solved in any incremental and efficient manner following the techniques in \cite{GraeserKornhuber2009,Tai2003}, namely by computations over hat function supports.\footnote{Compare equations (2.31) and (2.32) in \cite{Farrelletal2021}, which describe the same sense of locality.}  However, for general problems lacking the niceties of Example \ref{ex:fnice}, certain additional ideas are needed for practical implementation of CD-type iterations.  Section \ref{sec:vcycle} constructs a multiplicative multilevel CD iteration for a general operator $f:\cK\to\cV'$ by applying the full approximation storage (FAS) idea of Brandt \cite{Brandt1977}.

From now on we will restrict to the case of \emph{box constraints} \cite{BensonMunson2006,FerrisPang1997}, equivalently \emph{bi-obstacle problems}.  Suppose $\Omega \subset \RR^d$ is open and let $\mathcal{V}=W^{1,p}(\Omega)$.  The \emph{lower obstacle} $\underline{\gamma}$ and the \emph{upper obstacle} $\overline{\gamma}$ are given functions, defined on the closure $\overline{\Omega}$, with values in the extended real line $\tilde\RR = [-\infty,+\infty]$.  We suppose $\underline{\gamma} : \overline{\Omega} \to [-\infty,+\infty)$ and $\overline{\gamma} : \overline{\Omega} \to (-\infty,+\infty]$ are measurable, $\underline{\gamma} \le \overline{\gamma}$ on $\overline{\Omega}$, and $\underline{\gamma} \le g_D \le \overline{\gamma}$ on $\partial_D \Omega$, and that $\partial\Omega$ is split into disjoint, measurable sets, $\partial\Omega = \partial_D \Omega \cup \partial_N \Omega$, with $\partial_D \Omega$ of positive measure, to facilitate well-posedness \cite{Evans2010}, and $\partial_N \Omega$ possibly empty.  Dirichlet data\footnote{Dirichlet boundary values are in the trace sense \cite{Evans2010}, and $g_D$ is assumed to be as regular as needed.} $g_D:\partial_D \Omega \to \RR$ then defines
\begin{equation}
\cK = \left\{w\,:\,\underline{\gamma} \le w \le \overline{\gamma} \, \text{ and }\, w\big|_{\partial_D \Omega} = g_D\right\} \subset \cV =W^{1,p}(\Omega), \label{eq:originalconstraintset}
\end{equation}
the closed and convex \emph{constraint set}.  Note that we are treating inequality constraints and boundary conditions in a unified manner.\footnote{Any Neumann conditions over $\partial_N \Omega$ are determined by the operator formula for $f$, i.e.~by an integration over $\partial_N\Omega$.}

Recalling abstract VI problem \eqref{eq:vi}, from now on we only constraint sets \eqref{eq:originalconstraintset}.  Given continuous $f:\cK \to \cV'$, and $\ell \in \cV'$, we will assume that the following is well-posed for $u^*\in \cK$:
\begin{equation}
\ip{f(u^*)}{v-u^*} \ge \ip{\ell}{v-u^*} \qquad \text{for all } v\in \cK. \label{eq:boxdirichletvi}
\end{equation}
Note that $v-u^* \in W_0^{1,p}(\Omega)$.

@@

We seek to rapidly compute the approximate solution of box-constrained VI problems \eqref{eq:boxdirichletvi} using finite elements (FE).  Suppose $\Omega \subset \RR^d$ is a (open) bounded polygon.\footnote{Section \ref{sec:results} illustrates $d=1,2,3$ cases.}  We will construct a multilevel FE approximation to \eqref{eq:boxdirichletvi} over $\Omega$ using $P_1$ or $Q_1$ elements \cite{Elmanetal2014}, by refining a coarsest mesh into a mesh hierarchy; the goal is an accurate solution on the finest-level mesh.  The hierarchy allows the creation of a multilevel CD of $\cK$, using the ideas in the next two Sections, which permits solvers based on admissibility-preserving coarse corrections.

@@

In the VI material which follows, use of higher-order Lagrange elements, i.e.~$P_k$ or $Q_k$ with $k\ge 2$, would imply a ``variational crime'' \cite[Chapter 10]{BrennerScott2007} special to inequality constraints, which we avoid.  Namely, our FE construction of constraint sets (Section \ref{sec:cdmultilevel}) will require \emph{nodal monotonicity},
\begin{equation}
\bw \ge \bz \quad \implies \quad w \ge z \label{eq:nodalmonotonicity}
\end{equation}
for all $w,z \in \mathcal{V}^j$, where $\bw,\bz \in \RR^{m_j}$ denote vectors of nodal values.  However, for $k\ge 2$ there are hat functions in $P_k$ and $Q_k$, i.e.~functions with nodal values one or zero, which take on negative values \cite[Figure 1.7]{Elmanetal2014}.  Evidently \eqref{eq:nodalmonotonicity} is equivalent to the non-existence of such indefinite-sign hat functions.

To allow box constraints on all meshes in the hierarchy, let $\tilde{\mathcal{V}}^j$ denote the set of functions on the nodes of $\mathcal{T}^j$ with values in the extended real line $\tilde{\RR} = [-\infty,+\infty]$.  Suppose that the finest-level \emph{lower} and \emph{upper obstacles} $\underline{\gamma}^J, \overline{\gamma}^J \in \tilde{\mathcal{V}}^J$ satisfy
\begin{equation}
\underline{\gamma}^J \le \overline{\gamma}^J, \qquad \underline{\gamma}^J < +\infty, \qquad -\infty < \overline{\gamma}^J. \label{eq:fe:boxconstraintrequirements}
\end{equation}
Also assume that $\underline{\gamma}^J \le g_D^J \le \overline{\gamma}^J$ on those nodes of $\mathcal{T}^J$ which lie in $\partial_D \Omega$.  Define
\begin{equation}
\mathcal{K}^J = \left\{w\,:\,\underline{\gamma}^J \le w \le \overline{\gamma}^J \text{ on nodes of } \mathcal{T}^J, \, \text{ and } \, w|_{\partial_D\Omega} = g_D^J|_{\partial_D\Omega}\right\} \subset \mathcal{V}^J, \label{eq:fe:fineconstraintset}
\end{equation}
the finest-level \emph{constraint set} of admissible FE functions.  This is a nonempty, closed, and convex subset of $\mathcal{V}^J$.  As $\mathcal{V}^J$ functions never take values $\pm\infty$, inequality in \eqref{eq:fe:fineconstraintset} is strict where $\underline{\gamma}^J, \overline{\gamma}^J$ are infinite; there is no lower or upper constraint, respectively, at such nodes.

@@

While \eqref{eq:fe:vi} approximates \eqref{eq:boxdirichletvi}, we warn the reader that it may not technically be conforming for that VI.  That is, even if quadrature and other implementation details for $f^J$ are ignored, and despite the vector space inclusion $\mathcal{V}^J \subset \mathcal{V}$, we do not require or generally expect $\mathcal{K}^J \subset \mathcal{K}$.  Regardless of the smoothness of the continuum obstacles $\underline{\gamma}, \overline{\gamma}$, admissibility with respect to their FE interpolants $\underline{\gamma}^J, \overline{\gamma}^J$ will generally not imply continuum admissibility.\footnote{This caveat is especially relevant in geophysical situations where only nodal values of the obstacle are known.  Nodal elevation data generally misses the actual topographic detail in-between the nodes; compare \cite{Bueler2016}.}

The multilevel algorithm defined in Section \ref{sec:vcycle}, a solver for FE VI \eqref{eq:fe:vi}, requires transfer operators between meshes.  We require five such operators, of which three are standard \cite{Trottenbergetal2001}.  Table \ref{tab:transfers} will help the reader keep track.

@@

Canonical prolongation $P:\mathcal{V}^{j-1}\to\mathcal{V}^j$ is the vector space inclusion $\mathcal{V}^{j-1} \hookrightarrow \mathcal{V}^j$ of coarser-mesh functions.  Canonical (dual) restriction $R:(\mathcal{V}^j)'\to(\mathcal{V}^{j-1})'$ also acts as an identity in the sense that if $\sigma \in (\mathcal{V}^j)'$ then $(R\sigma)[z] = \sigma[z]$ for all $z \in \mathcal{V}^{j-1} \subset \mathcal{V}^j$.  These canonical transfers generally involve a change of FE representation.  For example, $R$ is ``full-weighting'' \cite{Trottenbergetal2001}.

To define injection $\iR:\mathcal{V}^j\to\mathcal{V}^{j-1}$, observe that the nodes of $\mathcal{T}^{j-1}$ are nodes of $\mathcal{T}^j$, and that for any values at $\mathcal{T}^j$ nodes there is a unique $\mathcal{V}^j$ function assuming those values.  For $z\in\mathcal{V}^j$ we define the injection $\iR z$ as the element of $\mathcal{V}^{j-1}$ with the same (coarse) nodal values.  Injection is monotonic: if $w \le z$ then $\iR w \le \iR z$.

The monotone restrictions $\maxR,\minR$ \cite{GraeserKornhuber2009}, which will only be applied to nodal functions $\tilde{\mathcal{V}}^j$, are nonlinear maps which maximize and minimize, respectively, nodal values over the open supports of coarser-mesh hat functions.  To make this precise, let $x_q^j$ denote the nodes of mesh $\mathcal{T}^j$, with corresponding hat functions $\psi_q^j(x)$.  For a $\tilde\RR$-valued nodal function $z\in\tilde{\mathcal{V}}^j$ we define functions in $\tilde{\mathcal{V}}^{j-1}$:
\begin{align}
(\maxR z)(x_p^{j-1}) &= \max \{z(x_q^j) \,:\, \psi_p^{j-1}(x_q^j) > 0\}, \label{eq:fe:monotonerestrictions} \\
(\minR z)(x_p^{j-1}) &= \min \{z(x_q^j) \,:\, \psi_p^{j-1}(x_q^j) > 0\}. \notag
\end{align}
For standard refinement a coarse-mesh node $x_p^{j-1}$ is in the set of fine nodes $x_q^j$ such that $\psi_p^{j-1}(x_q^j)>0$, so it follows that
\begin{equation}
\minR z \le z \le \maxR z,  \label{eq:fe:monotonerestrictionprops}
\end{equation}
On the other hand, $z\ge 0$ implies $\minR z \ge 0$ and $z \le 0$ implies $\maxR z \le 0$.  These ordering properties are used in the CD construction of the next two Sections.

@@

Our practical multilevel CD scheme uses \emph{defect constraints} \cite{GraeserKornhuber2009} which are based on an admissible finest-level iterate.  The next Example concretely illustrates why we do not apply restriction operators directly\footnote{An exception is in the F-cycle; see Section \ref{sec:implementation}.} to the finest-level box constraints $\underline{\gamma}^J,\overline{\gamma}^J$.  Our straightforward (iterate-dependent) bounds on the defect constraints, namely \eqref{eq:fe:chiordering} below, bypass the difficulties exposed in Example \ref{ex:directRbad}.

\begin{example}  \label{ex:directRbad}  
For concreteness, suppose $\Omega = [0,L] \subset \RR^1$ is an interval and consider a two-level mesh hierarchy ($J=1$).  Let $\underline{\gamma}^1$ be a sawtooth function with minimum value zero, and suppose $\overline{\gamma}^1=C+\underline{\gamma}^1$ with $C>0$ which is strictly less than the range of $\underline{\gamma}^1$.  Figure \ref{fig:directRbad} shows such functions; here levels $\mathcal{T}^1$ and $\mathcal{T}^0$ have $5$ and $3$ nodes, respectively.  Supposing empty Dirichlet boundary for simplicity, the constraint set $\mathcal{K}^1$ defined by \eqref{eq:fe:fineconstraintset} is nonempty.  However, various restriction schemes applied directly to $\underline{\gamma}^1,\overline{\gamma}^1$ are problematical.  Suppose we propose
    $$\underline{\gamma}^0, \overline{\gamma}^0 \equiv \iR \underline{\gamma}^1, \iR \overline{\gamma}^1. \qquad \text{\emph{[problematic]}}$$
Then $\underline{\gamma}^0=0$ and $\overline{\gamma}^0=C$ identically.  In that case the corresponding constraint set $\mathcal{K}^0$ is nonempty, but prolongation $Pw^0$ of an admissible solution of the coarse problem ($\underline{\gamma}^0 \le w^0 \le \overline{\gamma}^0$) is not admissible on the finer level: $Pw^0 \notin \mathcal{K}^1$.  On the other hand, suppose we propose
    $$\underline{\gamma}^0, \overline{\gamma}^0 \equiv \maxR \underline{\gamma}^1, \minR \overline{\gamma}^1. \qquad \text{\emph{[problematic]}}$$
The general motivation for this choice---compare \eqref{eq:fe:chilevels}---is that $w^0 \ge \underline{\gamma}^0 \, \implies \, Pw^0 \ge \underline{\gamma}^1$ and $w^0 \le \overline{\gamma}^0 \, \implies \, Pw^0 \le \overline{\gamma}^1$.  However, while these facts are true, the difficulty should be clear: the coarse-level admissible set $\mathcal{K}^0$ is now empty because the proposed lower bound $\underline{\gamma}^0$ actually exceeds the proposed upper bound $\overline{\gamma}^0$.
\end{example}

@@

Subtracting an admissible iterate from the original bounds provides a route out of the difficulties in Example \ref{ex:directRbad}.  Suppose $w^J \in \cK^J$ is an admissible finest-level iterate for \eqref{eq:fe:vi}.  We define the \emph{finest-level defect constraints} (DCs) \cite{GraeserKornhuber2009} as the following nodal functions in $\tilde{\mathcal{V}}^J$:
\begin{equation}
\underline{\chi}^J = \underline{\gamma}^J - w^J, \qquad \overline{\chi}^J = \overline{\gamma}^J - w^J. \label{eq:fe:defectconstraints}
\end{equation}
Definition \eqref{eq:fe:fineconstraintset} implies that $\underline{\chi}^J \le 0 \le \overline{\chi}^J$ everywhere in $\overline{\Omega}$, including on the Dirichlet boundary $\partial_D\Omega$.  If $\underline{\gamma}^J=-\infty$ then $\underline{\chi}^J=-\infty$ by definition, and similarly if $\overline{\gamma}^J=+\infty$ then $\overline{\chi}^J=+\infty$.  Observe that a corrected iterate $w^J + z^J$ is admissible, $w^J + z^J \in \cK^J$, if and only if the perturbation $z^J \in \mathcal{V}^J$ is zero on $\partial_D\Omega$ and bounded by the defect constraints, $\underline{\chi}^J \le z^J \le \overline{\chi}^J$.
